### 2009-12-07 docs/marcello-dissertation/thesis/chapter4-netbeams-in-depth-requirements-analysis-and-specification.tex
Most of the revisions and updates... Changes to chapters 3, 4 after Kleber's chat online (adding missing taxonomy).
Revisions of print chapters until chapter 5.

Chapter 7 is reviewed until before discussion. 

Next revision should include all the necessary reviews for the
release draft.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@676 559edde2-a4d4-44e0-96fc-4035e40d7396

\subsection{The SF-BEAMS Infrastructure}

@@

In general, the SF-BEAMS network infrastructure contains varying a number of
wired and wireless devices standing on pylons docked at the pier. Each of them
is responsible for observing different conditions of the area, having its own
mechanisms for internal storage for the collected data. In this way, data can
be directly transferred to the labs via Ethernet cables or collected manually
with a laptop computer.

\subsection{SF-BEAMS produced data and collection process}
\label{sec:sfbeams}

As described in section \ref{sec:sn-infrastructure}, sensor devices produce
its observed data based on properties of measurements defined by its
manufacturer. One example of such sensors is the ``YSI 6600 ESD V2", as seen in
Figure \ref{fig:ysi-device}. It is a powerful water quality monitoring device,
capable of producing around 52 bytes on a single real-time data stream reading,
as shown on Table \ref{tab:ysi-data-stream}.

@@

\begin{table}[!b]
    \caption{YSI Data Stream}
    \begin{center}
        \begin{tabular}{lr}
  12.20    192    179 5588.40   0.09   0.084   0.059  7.98   -79.6   99.5  8.83 
  0.4     8.7
        \end{tabular}
    \end{center}
    \label{tab:ysi-data-stream}
\end{table}

@@

By analyzing the maximum amount of produced data by the YSI sonde device is
around hundreds of Megabytes a year, as \textbf{483,840} samples are generated
by one single device. Upon collecting data from sensors, the RTC staff uses
automation scripts written in Matlab \cite{matlab} to process, index and
distribute the raw data in different formats. One of these formats is the
OPeNDAP \cite{opendap}, which is widely used at research institutions that
promote easier data exchange among them. The SF-BEAMS website provides access
to the collected data through the Internet at
http://sfbeams.sfsu.edu:8080/opendap. An example of an HTTP Request from
accessing the data for a specific YSI is shown in Listing
\ref{file:rtc-ysi-opendap}.

@@

\begin{itemize}
  \item \textbf{The Purpose of Sensor Data}: the data from SF-BEAMS is solely
  used for the purpose of Data Archival, since they are stored in the RTC's
  lab;
  \item \textbf{The Location of the Sensor Data}: it is clear that the purpose
  of data is directly related to its location and, as a consequence, the
  strategy of \textbf{External Data Storage} is used to store its collected
  data at the network sink at the RTC's labs;
  \item \textbf{Data Model}: since the format uses by the RTC staff to share
  data among researcher is in the OPEnDAP format, it is clear that the data
  model used is the tabular data model, since it uses the comma-delimited
  files. In this way, the \textbf{Schema-less} is used;
  \item \textbf{Data Provenance}: once the collected data reaches the RTC
  lab, a quality assurance process is executed, and the data is transformed
  into the format used by OPEnDAP. This conversion carries data regarding
  the time of the data collection, and therefore, the use of one of the
  \textbf{Time Dimensions} such as the valid time. Since the files are stored
  with specific file name structure, using timestamps to describe the data, the
  use of transaction time may be also considered. Finally, the contents of the
  files contain the \textbf{Data Identity} as shown in the samples of
  the collected in Listing \ref{file:rtc-ysi-opendap};
  \item \textbf{Query Processing Mechanism}: similar to the data stored in a
  centralized way, the query processing is also \textbf{Centralized}, as the
  data is stored on file system;
  \item \textbf{Database System Organization}: there are no databases in use,
  but the use the OPeNDAP Hyrax server as a middleware that servers the
  collected data using the Internet.
\end{itemize}

@@

The scope of this work then is defined as to provide a persistence layer for
NetBEAMS, making a selection of a persistence layer that is an exception to the
taxonomical classification of SF-BEAMS. Furthermore, the technology to be
selected for the persistence layer must take into account the requirements of
NetBEAMS primary users such as Computer Science researchers, as well as end
users such as Biologists from the RTC laboratories. The requirements for the
data persistece specifically for NetBEAMS is described in the following section.

>>>>

\subsection{The SF-BEAMS Infrastructure}
\label{sec:sfbeams}

@@

In general, the SF-BEAMS network infrastructure contains varying a number of
wired and wireless devices standing on pylons docked at the pier. Each of them
is responsible for observing different conditions of the area, having its own
mechanisms for internal storage for the collected data. In this way, data can
be directly transferred to the labs via Ethernet cables or collected manually
with a laptop computer. The current data collection process for SF-BEAMS is
described in figure
\ref{fig:SF-BEAMS-system-architecture}\cite{sfbeams-current-system}. Upon
collecting data from sensors, the RTC staff uses automation scripts written in
Matlab \cite{matlab} to process, index and distribute the raw data in
different formats. One of these formats is the OPeNDAP \cite{opendap}, which
is widely used at research institutions that promote easier data exchange
among them. The SF-BEAMS website provides access to the collected data through
the Internet at http://sfbeams.sfsu.edu:8080/opendap. An example of an HTTP
Request from accessing the data for a specific YSI is shown in Listing
\ref{file:rtc-ysi-opendap}.

\begin{figure}[!b]
  \centering
  \includegraphics[scale=0.5]{../diagrams/SF-BEAMS-system-architecture}
  \caption{The Current SF-BEAMS Data Collection Process}
  \label{fig:SF-BEAMS-system-architecture}
\end{figure}

As described in section \ref{sec:sn-infrastructure}, sensor devices produce
its observed data based on properties of measurements defined by its
manufacturer. NetBEAMS has used ``YSI 6600 ESD V2", as seen in Figure
\ref{fig:ysi-device}, one of the sensors devices used by SF-BEAMS. It is a
powerful water quality monitoring device, capable of producing around 52 bytes
on a single real-time data stream reading, as shown on Table
\ref{tab:ysi-data-stream}.

@@

\begin{table}
    \begin{center}
        \begin{tabular}{|l|}\hline
  12.20    192    179 5588.40   0.09   0.084   0.059  7.98   -79.6   99.5  8.83  0.4     8.7\\\hline
        \end{tabular}
    \end{center}
    \label{tab:ysi-data-stream}
    \caption{The Collected Data Stream from the YSI Sonde 6600ESDV2}
\end{table}

@@

By analyzing the maximum amount of produced data by the YSI sonde device is
around hundreds of Megabytes a year, as \textbf{483,840} samples are generated
by one single device.

@@

\begin{itemize}
  \item \textbf{The Purpose of Sensor Data}: the data from SF-BEAMS is solely
  used for the purpose of Data Archival, since they are stored in the RTC's
  lab;
  \item \textbf{The Location of the Sensor Data}: it is clear that the purpose
  of data is directly related to its location and, as a consequence, the
  strategy of \textbf{External Data Storage} is used to store its collected
  data at the network sink at the RTC's labs;
  \item \textbf{Data Model}: since the format uses by the RTC staff to share
  data among researcher is in the OPEnDAP format, it is clear that the data
  model used is the tabular data model, since it uses the comma-delimited
  files. In this way, the \textbf{Schema-less} is used;
  \item \textbf{Data Description}: once the collected data reaches the RTC
  lab, a quality assurance process is executed, and the data is transformed
  into the format used by OPEnDAP. This conversion carries data regarding
  the time of the data collection, and therefore, the use of one of the
  \textbf{Time Dimensions} such as the valid time. Since the files are stored
  with specific file name structure, using timestamps to describe the data, the
  use of transaction time may be also considered. Finally, the contents of the
  files contain the \textbf{Data Identity} as shown in the samples of
  the collected in Listing \ref{file:rtc-ysi-opendap};
  \item \textbf{Query Processing Mechanism}: similar to the data stored in a
  centralized way, the query processing is also \textbf{Centralized}, as the
  data is stored on file system;
  \item \textbf{Data Volume}: the volume of data produced is in the range of
  small. It is related to the volume of data produced by the sensor devices,
  which is numerical data streams, and the total number of sensor devices;
  \item \textbf{System Organization}: SF-BEAMS runs on a Single System, as
  the collected data is stored directly in the file-system after previous 
  processing. However, they provide the collected data to the Internet using
  the OPeNDAP Hyrax system as a middleware.
\end{itemize}

@@

The scope of this work is defined as to provide a persistence layer for
NetBEAMS, making a selection of a persistence layer that is an exception to the
taxonomical classification of SF-BEAMS. Furthermore, the technology to be
selected for the persistence layer must take into account the requirements of
NetBEAMS primary users such as Computer Science researchers, as well as end
users such as Biologists from the RTC laboratories. The requirements for the
data persistece specifically for NetBEAMS is described in the following section.

### 2009-12-07 docs/marcello-dissertation/thesis/chapter5-taxonomic-persistence-storage-selection-for-netbeams.tex
Most of the revisions and updates... Changes to chapters 3, 4 after Kleber's chat online (adding missing taxonomy).
Revisions of print chapters until chapter 5.

Chapter 7 is reviewed until before discussion. 

Next revision should include all the necessary reviews for the
release draft.

git-svn-id: svn://localhost/netbeams/branches/marcello/persistence@676 559edde2-a4d4-44e0-96fc-4035e40d7396

\chapter{Data Persistence Technology for Sensor Networks: An Empirical-based
Choice for NetBEAMS}

This chapter analyzes each of the taxonomies defined in chapter
\ref{chap:taxonomies} against different other variables. First, the
requirements defined by NetBEAMS, as our case study in chapter
\ref{sec:problem-requirements}, is to be compared. Then, the scope of this
work is compared to the technologies selected based on the readings of the
literature reviewing data persistence in sensor networks and other recent
trends in performance. Finally, a table summarizes the contenders strengths
and discusses the selected technology to be integrated with the design of a
Data Persistence component.

@@

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: an open-source relational database system
  that is the primary choice of the development of projects of different sizes.
  It was selected because of its popularity in the contemporary community;
  \item \textbf{Oracle}\cite{oracle}: is the most popular commercial database
  system, having a campus wide academic licenses available;
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks
  implementation \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database
  system described in the article that questions the use of the relational
  model \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. It was selected because NetBEAMS can
  represent the collected data in the XML format as shown in setion 
  \ref{sec:dsp-message}.
\end{itemize}

@@

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-table-partition, db-partitioning-relational}. Both former and the
latter are used in regular schema-dependent models or schema-less models,
which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas, db-partitioning-relational,
db-partitioning-relational-oracle}.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.65]{../diagrams/database-sharding-by-region}
  \caption{Example of Data-Centric Storage for Sensor Networks using Database
  Sharding}
  \label{fig:database-sharding-by-region}
\end{figure}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports External or Data-Centric Storage approaches.
  When Data-Centric Storage is used, both vertical and horizontal partitions
  \cite{db-partitioning-relational} can be  chosen, using different strategies
  to partition the data;
  \item \textbf{Oracle}: Supports External or Data-Centric Storage approaches.
  Only sup-  ports vertical partitioning with the definition of different types
  \cite{db-table-partition, db-partitioning-relational-oracle} when using
  Data-Centric storage;
  \item \textbf{Berkeley TinyDB}: Only supports the External Storage;
  \item \textbf{MongoDB}: Supports External or Data-Centric Storage. When the
  latter approach is used, the horizontal partition strategy is available with
  automatic shards \cite{db-mongo-partition};
  \item \textbf{DB2}: Supports External and Data-Centric storage. Only supports
  vertical table partition \cite{db-db2-partition} for the data-centric storage.
\end{itemize}

@@

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.65]{../diagrams/Relational-Model-Original}
  \caption{Relational Data Model for NetBEAMS - A first prototype}
  \label{fig:Relational-Model-Original}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.65]{../diagrams/Relational-Model-Addition-Modified}
  \caption{Relational Data Model for NetBEAMS - Modified version with new
  entity}
  \label{fig:Relational-Model-Addition-Modified}
\end{figure}

Although this data model have been used in different types of applications,
developers have tried to use the Relational Model to model an application that
could prevent any database change through the use of a technique that relates a
key to a value, called Key-Value data model. With the advent of Internet
applications, the need of such a model that does not need constant refactoring
led developers to propose such approach using the relational model as
documented in technical articles such as  and
. Based in these articles, consider the relational
model depicted in Figure  as a data model for
persisting data to our case study using the key-value strategy. The problem
with it is that key repetition occurs, since the nature of time-series data
requires the use of a timestamp key for each property of the sensor. Therefore,
this strategy is not well-suited to provide persistence data from collected
sensor networks.

There are no records of the use of this data model with Sensor Networks.
Different variants of such data model is the document-oriented data model,
where data is modeled using structured documents. However, these documents have
a dynamic structure that can be freely described without the use of a schema
validation mechanism as used in XML Schema. One example of such approach is the
use of the JSON data format [Ste08] in the implementation of database systems
which are becoming standard with the latest trend called Cloud Computing [LKN+
09]. Databases implementing this strategy is the mongoDB \cite{mongodb},
CouchDB \cite{couchdb}, Amazon Dynamo \cite{db-kvp}, among others.

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describe the data, depicted in section
\ref{sec:sn-provenance}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.75]{../diagrams/persistence-example-kvp}
  \caption{KVP Database instance Prototype}
  \label{fig:persistence-example-kvp}
\end{figure}

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and can be summarized as follows:

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships;
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic operations CRUD\footnote{Create-Retrieve-Update-Update
      database operations} data are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, contains functions for aggregation and complex filter.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via API\footnote{Application
      Programming Interface} through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like =, !=, <, > that ca be applied;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC\footnote{Open Database
      Connectivity};
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Systems tend to provide SOAP/REST \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{itemize}
  \item MySQL: supports the Relational Model;
  \item Oracle: supports Relational Model;
  \item Berkeley TinyDB: supports the Relational Model; 
  \item \textbf{MongoDB}: supports Document-Oriented Model;
  \item IBM DB2: supports the Relational Model and the XML model.
\end{itemize}

\section{Analysis of the Database System Organization Taxonomy}

@@

\subsection{Advantages of Database Partitioning and Sharding}

@@

\subsection{Disadvantages of Database Partitioning and Sharding}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports Data Replication;
  \item \textbf{Oracle}: Supports Data Replication;
  \item \textbf{Berkeley TinyDB}: NO Supports to distributed data;
  \item \textbf{MongoDB}: Supports Data Replication;
  \item \textbf{IBM DB2}: Supports Data Replication.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Is  open-source, supports hot backup and contains
  scripting APIs from  the community;
  \item \textbf{Oracle}: It is not Open-Source, but contains lots of support
  from the community. Also APIS are available for the majority of languages;
  \item \textbf{Berkeley TinyDB}: Not open-source;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from  the community, offering native APIS in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not Open-Source, but offers support from its
  community.
\end{itemize}

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \caption{Amount of data produced by the RTC's YSI sondes}
        \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{Oracle} &
        \textbf{TinyDB} & \textbf{MongoDB} & \textbf{IBM DB2}\\\hline 
        Centralized Query & + & + & + & + & + \\\hline 
        Distributed System & ++ & ++ & - & ++ & +\\\hline 
        Schema-less & - & - & - & + & +\\\hline 
        Provenance Support & + & + & + & + & +\\\hline 
        NO-SQL Query & - & - & - & + & +\\\hline 
        Data Partition & + & + & - & + & +\\\hline 
        Hot Backup & + & + & - & - & +\\\hline 
        Export Capability & + & + & - & + & -\\\hline 
        Programming Lang. & + & + & + & + & +\\\hline
        Script Lang. & + & + & - & + & +\\\hline
        Open-Source & + & - & + & + & -\\\hline
        \end{tabular}
        \end{center}
\end{table}

>>>>

\chapter{Empirical Analysis of Data Persistence Taxonomies for NetBEAMS}

Providing data persistence for an existing sensor networks require analyzes of
the infrastructure and the nature of the produced data. Based on the taxonomies
of chapter 3, and the requirements of a data persistence for NetBEAMS in
chapter 4, this chapter aims at analyzing a group of database technologies
selected from literature review topics covered in chapter 2. Similarly,
supported by the scope of the required implementation presented in this
chapter, the comparison of the technogies is summarized, highlighting the
selected database system used for the design of a component for NetBEAMS and
used for the design of the experiments.

@@

\begin{itemize}
  \item \textbf{MySQL}\cite{mysql}: system used in the implementation of a
  dateware house with sensor network data \cite{sn-dataware-house};
  \item \textbf{TinyDB}\cite{tinydb}: as referred in the literature review,
  TinyDB is a database system used in many different sensor networks
  implementation \cite{sn-db-tinydb};
  \item \textbf{MongoDB}\cite{mongodb}: an open-source key-value database
  system described in the article that questions the use of the relational
  model \cite{db-is-rdbs-dommed}, as well as the Cloud Computing trends survey
  \cite{cloud-comp-architectures};
  \item \textbf{DB2}\cite{db2}: in the pursuit of an XML database system, DB2
  is listed as a hybrid database system that offers both the relational and the
  XML models \cite{db-xml-enabled}. It was selected because NetBEAMS can
  represent the collected data in the XML format as shown in setion 
  \ref{sec:dsp-message}.
\end{itemize}

@@

The horizontal or vertical partitioning strategies are also referred to as
Database Sharding \cite{db-shard-discussion}, or simply Database partitioning
\cite{db-table-partition, db-partitioning-relational}. Both former and the
latter are used in regular schema-dependent models or schema-less models,
which uses denormalized data models. There are different approaches to
partition data \cite{db-shard-schemas, db-partitioning-relational}.

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/database-sharding-by-region}
  \caption{Example of Data-Centric Storage for Sensor Networks using Database
  Sharding}
  \label{fig:database-sharding-by-region}
\end{figure}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports External or Data-Centric Storage approaches.
  When Data-Centric Storage is used, both vertical and horizontal partitions
  \cite{db-partitioning-relational} can be chosen, using different strategies
  to partition the data;
  \item \textbf{TinyDB}: Only supports the External Storage;
  \item \textbf{MongoDB}: Supports External or Data-Centric Storage. When the
  latter approach is used, the horizontal partition strategy is available with
  automatic shards \cite{db-mongo-partition};
  \item \textbf{DB2}: Supports External and Data-Centric storage. Only supports
  vertical table partition \cite{db-db2-partition} for the data-centric storage.
\end{itemize}

@@

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/Relational-Model-Original}
  \caption{Relational Data Model for NetBEAMS - A first prototype}
  \label{fig:Relational-Model-Original}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[scale=0.5]{../diagrams/Relational-Model-Addition-Modified}
  \caption{Relational Data Model for NetBEAMS - Modified version with new
  entity}
  \label{fig:Relational-Model-Addition-Modified}
\end{figure}

For instance, all data needed for the YSI sonde data, as well as all necessary
provenance data that describe the data, depicted in section
\ref{sec:sn-data-description}, are represented by means of key-value pairs
\ref{fig:persistence-example-kvp}.

\begin{figure}[!h]
  \centering
  \includegraphics[scale=0.5]{../diagrams/persistence-example-kvp}
  \caption{KVP Database instance Prototype}
  \label{fig:persistence-example-kvp}
\end{figure}

The relational and key-value data models were compared by
\cite{db-is-rdbs-dommed} and is compared on table
\ref{tab:schema-vs-schemaless}:

\begin{table}
    \label{tab:schema-vs-schemaless}
    \caption{Schema-Dependent X Schema-less Databases Compared: Properties}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Real-world model represented by entities, classified in tables;
      \item Tables composed by columns and rows. Rows are comprised by column
      values, which have the same schema;
      \item Data Model defined in advance with a schema, which contains
      relationships and constraints to enforce data integrity;
      \item Data represents ``natural'' entities, not application specific;
      \item Data Normalization is the data structuring process to remove data
      duplication, as well as establishing data associations through table
      relationships;
      \item Data Provenance can be provided through data types such as
      ``Timestamp'' for time or float for the GPS float-based coordinates; 
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Real-world model by entities, classified in Domains;
      \item Domains are similar to tables, but like buckets that contains items
      without a pre-defined schema, what enable them to have different schemas;
      \item Items are identified by keys, as well as have a dynamic set of
      attributes attached to it, however with no schema defined;
      \item Items may represent not only the natural representation of data, but
      also include application-specific data;
      \item Since data may be repeated, no data normalization is done, so that
      integrity is done in the application layer;
      \item Data provenance can be added through the use of keys and relating
      values for the time information and location of the data.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:schema-vs-schemaless-daccess}
    \caption{Schema-Dependent X Schema-less Databases: Data Access}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item The basic operations CRUD\footnote{Create-Retrieve-Update-Update
      database operations} data are performed using a structured language such
      as the SQL or XPath;
      \item Query languages can access data from different tables through
      joins, contains functions for aggregation and complex filter.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item The CRUD operations are performed via API\footnote{Application
      Programming Interface} through programming languages;
      \item Some technologies provide basic SQL-like syntax for filter criteria
      with some predicates like =, !=, <, > that ca be applied;
      \item The data and application integrity logic is placed in the
      application layer.
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{table}
    \label{tab:tab:schema-vs-schemaless-api-interface}
    \caption{Schema-Dependent X Schema-less Databases: Application Interface}
    \begin{center}
    \begin{tabular}{|p{210pt}|p{210pt}|}\hline
    Schema-Dependent Databases & Schema-less Databases\\\hline
    \begin{enumerate}
      \item Have their own specific API or through ODBC\footnote{Open Database
      Connectivity};
      \item Data is stored in a format that represents its natural structure,
      and for this reason, in a single or distributed fashion.
    \end{enumerate} 
    & 
    \begin{enumerate}
      \item Systems tend to provide SOAP/REST \cite{http-rest};
      \item \cite{db-is-rdbs-dommed} claims that data is stored in a more
      effective way, requiring only code plumbing for the relational code;
    \end{enumerate}
    \\\hline
    \end{tabular}
    \end{center}
\end{table}

\begin{itemize}
  \item \textbf{MySQL}: supports the Relational Model;
  \item \textbf{TinyDB}: supports the Relational Model; 
  \item \textbf{MongoDB}: supports Document-Oriented Model;
  \item \textbf{IBM DB2}: supports the Relational Model and the XML model.
\end{itemize}

@@

\section{Analysis of the System Organization Taxonomy}

@@

\subsection{Advantages of Database Partitioning}

@@

\subsection{Disadvantages of Database Partitioning}

@@

\begin{itemize}
  \item \textbf{MySQL}: Supports Data Replication;
  \item \textbf{TinyDB}: NO Supports to distributed data;
  \item \textbf{MongoDB}: Supports Data Replication;
  \item \textbf{IBM DB2}: Supports Data Replication.
\end{itemize}

@@

\begin{itemize}
  \item \textbf{MySQL}: Is  open-source, supports hot backup and contains
  scripting APIs from  the community;
  from the community. Also APIS are available for the majority of languages;
  \item \textbf{TinyDB}: Not open-source;
  \item \textbf{MongoDB}: An open-source database system with an increasing
  support from  the community, offering native APIS in different programming
  languages and scripting languages;
  \item \textbf{IBM DB2}: Not Open-Source, but offers support from its
  community.
\end{itemize}

@@

\begin{table}
    \label{tab:ysi-data-distribution}
    \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|}\hline 
        \textbf{Database} & \textbf{MySQL} & \textbf{TinyDB} & \textbf{MongoDB} & \textbf{IBM DB2}\\\hline
        Centralized Query & + & + & + & + \\\hline 
        Distributed System & ++ & - & ++ & +\\\hline 
        Schema-less & - & - & + & -\\\hline 
        Provenance Support & + & + & + & +\\\hline 
        NO-SQL Query & - & - & + & -\\\hline 
        Partition & + & - & + & +\\\hline 
        Hot Backup & + & - & + & +\\\hline 
        Export Capability & + & - & + & -\\\hline 
        Programming Lang. & + & + & + & +\\\hline
        Script Lang. & + & - & + & +\\\hline
        Open-Source & + & + & + & -\\\hline
        \end{tabular}
    \end{center}
\end{table}
